A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-10-09 latex/main.tex
Pass through first para of Results

\author[1,6]{Shing~H.~Zhan}
% TODO add joint second here?
\author[1,5]{Yan~Wong}
\author[2,5]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1]{Isobel~Guthrie}
\author[1]{Benjamin~Jeffery}
\author[1]{Duncan~S.~Palmer}
\author[3]{Carmen~Lia~Murall}
\author[4]{Sarah~P.~Otto}
\author[1\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

Using sc2ts, we built an ARG from 2,482,157 samples
from the Viridian v04 dataset \cite{Hunt2024},
collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a server with 64 cores (2X AMD EPYC 7502)
and 512 GB of RAM ($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG
contains 2,689,054 nodes,  % Note the estimated dates ?
2,748,838 edges, and 2,285,344 mutations at 29,893 sites.
Including sample IDs and the Pango and Scorpio designations computed by
Pangolin~\cite{OToole2021assignment} as metadata for each node,
the ARG is stored in a 32MiB tszip format file.
% It's confusing to talk about recombinant QC here as we're talking 
% about the full ARG for the rest of the section.
Of the nodes, 855 represent recombination events,
resulting in 316 distinct trees along the genome
(some events occurred at the same breakpoints [vague, rephrase]).
[TODO note about loading up the time it takes to get the node 
data dataframes in a notebook?]

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination intervals.
(A) Distribution of the breakpoint intervals of 354 recombination events 
along the genome (annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) FIXME Heatmap showing the co-occurrence counts of the parents classified by Scorpio designations.
(D) Relationship between breakpoint interval size and time between the recombination event
and the parental MRCA. The black curve shows the theoretical expected relationship
between these quantities, estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
\label{fig:recombinant_intervals}
\end{figure}

>>>>

\author[1,6]{Shing~H.~Zhan}
\author[1,5]{Yan~Wong}
\author[2,5]{Anastasia~Ignatieva}
\author[3]{Katherine~Eaton}
\author[1]{Isobel~Guthrie}
\author[1]{Benjamin~Jeffery}
\author[1]{Duncan~S.~Palmer}
\author[3]{Carmen~Lia~Murall}
\author[4]{Sarah~P.~Otto}
\author[1\orcidlink{0000-0002-7894-5253}]{Jerome~Kelleher}

@@

Using sc2ts, we built an ARG from 2,482,157 samples
from the Viridian v04 dataset \cite{Hunt2024},
collected between 2020-01-01 and 2023-02-20.
Inference took 21 days on a server with 64 cores (2X AMD EPYC 7502)
and 512 GB of RAM ($\sim 5$ CPU years; Figure~\ref{fig:inference_resources}).
After postprocessing, the ARG
contains 2,689,054 nodes,  % Note the estimated dates ?
2,748,838 edges, and 2,285,344 mutations at 29,893 sites,
and is stored in a 32MiB file.
Of the nodes, 855 represent recombination events,
resulting in 316 distinct trees along the genome
(some events occurred at the same breakpoints [vague, rephrase]).
% NOTE: if we keep this we should redo on Yan's laptop and quote those numbers
Using the tskit Python API, the ARG can be loaded into 
memory in less than a second [Concrete numbers?], and the 
vectorised representation integrates tightly with 
powerful data-science tools such as NumPy~\cite{Harris2020array}.
For example, a tabular summary for the 2.7m nodes, including
sample IDs and Pango designations (stored as metadata in the ARG),
can be derived as a Pandas dataframe~\cite{Mckinney2010data} in 0.6s.
The ARG can therefore be easily and efficiently analysed using familiar tools
with minimal hardware requirements.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{recombinant_intervals}
\caption{
Recombination breakpoint intervals.
(A) Distribution of the breakpoint intervals of
354 recombination events along the genome
(annotated with gene labels on top).
Each line represents an interval colored by its size.
(B) Number of overlapping breakpoint intervals.
(C) Distribution of breakpoint interval size (bases).
(D) Relationship between breakpoint interval size and
time between the recombination event and the parental MRCA.
The black curve shows the theoretical expected relationship
between these quantities,
estimated by minimizing the sum of squared residuals
given a single parameter (the mutation rate) (STAR Methods).}
(E) Distribution of time interval size (months).
\label{fig:recombinant_intervals}
\end{figure}
